---
output: 
  github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please only edit the Rmd file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/vignette-supp/",
  echo=TRUE, 
  warning=FALSE, 
  message=FALSE,
  tidy=TRUE
)
```

<!-- badges: start -->
[![R-universe version](https://sudaraka88.r-universe.dev/strainspy/badges/version)](https://sudaraka88.r-universe.dev/strainspy)
[![R-universe checks](https://sudaraka88.r-universe.dev/strainspy/badges/checks)](https://sudaraka88.r-universe.dev/strainspy#checktable)
<!-- Download counts are not available from R-universe. Uncomment these once
     the package is published on bioconda and accepted into Bioconductor:
[![bioconda downloads](https://img.shields.io/conda/dn/bioconda/r-strainspy.svg)](https://anaconda.org/bioconda/r-strainspy)
[![Bioconductor downloads](https://bioconductor.org/shields/downloads/release/strainspy.svg)](https://bioconductor.org/packages/strainspy)
-->
[![Databases](https://zenodo.org/badge/DOI/10.5281/zenodo.21796878.svg)](https://zenodo.org/records/21796878)
<!-- badges: end -->


# Strainspy

## Installation

`StrainSpy` is currently available on github. It can be installed with `remotes`

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("gtonkinhill/strainspy")
```


If you would like to also build the vignette with your installation run:
```{r, eval=FALSE}
remotes::install_github("gtonkinhill/strainspy", build_vignettes = TRUE)
```

### Alternative: install from R-universe

`StrainSpy` is also published on
[R-universe](https://sudaraka88.r-universe.dev/strainspy)

```{r, eval = FALSE}
install.packages("BiocManager")
install.packages("strainspy",
                 repos = c("https://sudaraka88.r-universe.dev",
                           BiocManager::repositories()))
```

## Analysing data with StrainSpy

A full set of examples and analyses performed using `StrainSpy` are available [here](https://sudaraka88.github.io/strainspy-manuscript/). Raw R scripts used in the analysis are available at: https://github.com/Sudaraka88/strainspy-manuscript.

Several useful notes for using StrainSpy with your own data is provided at the [end of this read me](#analysing-your-own-data).

## Quick Start

This quick start demonstrates a typical `StrainSpy` analysis and showcases some of the models and outputs available. Here, we analyse a 200 sample subset of the data described in [Wallen *et al.* 2022](https://doi.org/10.1038/s41467-022-34667-x). 

```{r, fig.width =8, fig.height=6, fig.align='center'}
library(strainspy)
library(ggplot2)

example_meta_path <- system.file("extdata", "example_metadata.csv.gz", package = "strainspy")
example_sylph_path <- system.file("extdata", "example_sylph_profile.tsv.gz", package = "strainspy")
example_taxonomy_path <- system.file("extdata", "example_taxonomy.tsv.gz", package = "strainspy")
```

**NOTE:** When analysing your own data, be sure to replace these example paths with valid file paths on your system. The `system.file()` function is used here only for demonstration purposes in this vignette.

### Load and filter data

```{r}
# Read in metadata
meta_data <- readr::read_csv(example_meta_path)

meta_data$Case_status = factor(meta_data$Case_status) # Required for visualising
# Read in sylph profile
se <- read_sylph(example_sylph_path, meta_data =  meta_data)

# Filter by presence. This will remove any strains that are not present in at least 30 samples
se <- filter_by_presence(se, min_nonzero = 30)

# Read in taxonomy
taxonomy <- read_taxonomy(example_taxonomy_path)
```

### Fit the model

```{r}
# Create design matrix
design <- as.formula(" ~ Case_status + Age_at_collection")

# Fit a Zero-inflated beta model using the default preset_weak MAP prior
fit <- glmZiBFit(se,  design, nthreads=parallel::detectCores())
```

### Summarise and plot the results

```{r, fig.width=8, fig.asp=0.75, out.width='100%'}
# Get top hits
th = top_hits(fit, coef = 2)

# Create Volcano plot
plot_volcano(fit, label = T)

knitr::kable(th)
```

*Coprobacillus cateniformis* shows difference in presence (adjusted p = 0.0213). *Clostridiales bacterium* shows difference in identity (adjusted p < 0.01). 

### Perform post-hoc testing to validate the beta hit

```{r}
th = comp_ani_diff_and_posthoc_test(se, fit, th)

# This NA indicates posthoc testing does not invalidate this beta hit
is.na(th$Comment)

th$ANI_Difference
```

In *Clostridiales bacterium*, average ANI (green dashed line) is 0.25% lower in Controls compared to Parkinson Disease (PD). The distribution appears different between groups.

## Visualise the distribution of top hits with Case_status

```{r, fig.width=8, fig.asp=0.75, out.width='100%'}
plot_ani_dist(se, 'Case_status', top_hits(fit)$Contig_name)
plot_histogram(se, 'Case_status', top_hits(fit)$Contig_name[2], 
               drop_zeros = TRUE, fit_spline = T, bins = 54, separate_facets = T)
```

## Incorporate taxonomy

### Generate taxonomy informed Manhattan plot with adjusted p-values
```{r, fig.width=8, fig.asp=0.75, out.width='100%'}
plot_manhattan(fit, taxonomy = taxonomy)
```

### Create a traditional Manhattan plot coloured by taxonomy with unadjusted p-values and Bonferroni significance thresholds
```{r, fig.width=10, fig.height=6}
plot_manhattan(fit, taxonomy = taxonomy, aggregate_by_taxa = F)
```

## Estimate sample richness

This can be used as a proxy for alpha diversity. 

```{r, fig.width=8, fig.asp=0.75, out.width='100%'}
rchns = estimate_sample_richness(se, taxonomy)$species_richness
df_rch = merge(
  data.frame(run_accession = names(rchns), value = rchns),
  meta_data,
  by = "run_accession"
)

ggplot(df_rch, aes(Case_status, value, fill = Case_status)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  theme_classic() + ylab("Number of Species")

```

## Estimate between sample diversity

```{r, fig.width=8, fig.asp=0.75, out.width='100%', warning=FALSE, message=FALSE, results='hide'}
divs = strainspy::estimate_beta_diversity(se, phenotype = 'Case_status', return_plots = T) 

divs$plots$NMDS
```

## Working with other inputs

### Example using Sourmash output
```{r}
example_sourmash_path <- system.file("extdata", "example_sourmash.csv.gz", package = "strainspy")
sm <- read_sourmash(example_sourmash_path, meta_data)
```
All remaining functions are compatible with this output. 

**Note:** `strainspy` provides a function to merge `sourmash gather` and `sourmash search` outputs. See help for details.

### Example using MetaPhlAn output
```{r}
example_metaphlan_path <- system.file("extdata", "metaphlan_merged.tsv.gz", package = "strainspy")
example_taxonomy_path <- system.file("extdata", "metaphlan_taxonomy.tsv.gz", package = "strainspy")
mp <- read_metaphlan(example_metaphlan_path, meta_data)
```
All remaining functions are compatible with this output. 

**Note:** `strainspy` provides a function to merge `MetaPhlAn` profiles and generate the taxonomy file. See help for details.

# Analysing your own data

To analyse your own metagenomic data with StrainSpy, you first need to profile your metagenomes.

### Profiling using Sylph

We recommend using `Sylph query` with the [99% GTDB](https://zenodo.org/records/21796878) database for strain-level analysis. 

See the [Sylph documentation](https://sylph-docs.github.io/install%2Bquickstart/) for full installation details.

For paired-end reads, a typical command is:

```
sylph query "$db_path" -u --read-seq-id 99.9 -t "$num_cpu_threads$ -1 "$R1" -2 "$R2" -o "$out.tsv"
```

For each metagenome, you need to provide the input path to the sylph database (`$db_path`), the number of threads to use (`$num_cpu_threads`), paths to forward and reverse reads (`$R1` and `$R2`), and the output path for the results (`$out.tsv`). For single-end read data, use:

```
sylph query "$db_path" -u --read-seq-id 99.9 -t "$num_cpu_threads$ -r "$reads" -o "$out.tsv"
```

This will create a unique output `tsv` file for each metagenome.

### Combining Sylph outputs

`StrainSpy` can directly accept a vector of paths to individual Sylph outputs into `read_sylph()`. Internally, this will call the function `merge_sylph_files()` to combine these into a single temp file. Alternatively, you can call `merge_sylph_files()` to combine Sylph outputs and save them to a single file that can also be used with `read_sylph()`, similar to the example in [Quick Start](#quick-start).

**Tip:** Make sure that all output files were generated using the same Sylph database and analysis settings (i.e., `query` or `profile`).

### Running Sylph on an HPC cluster

Running Sylph separately on hundreds or thousands of metagenomes can be cumbersome, particularly when the metagenomic reads need to be downloaded from a public repository. Several useful bash scripts are available [here](https://github.com/Sudaraka88/slurm_scripts/tree/main/sylph_download_run_detele) to download and use. These are written to automatically download, run Sylph, and delete the metagenomic reads, followed by merging Sylph outputs to a single tsv file ready for `StrainSpy`.

### Taxonomy data

Relevant taxonomy `tsv` files for each database is available for download at [Zenodo](https://zenodo.org/records/21796878).

### Metadata

A metadata `tsv` file (optionally `gz` compressed) needs to be prepared for each dataset before analysis. Each sample should be in a single row, and each column should contain a metadata variable. The first column should be named `run_accession` and contain the unique sample identifiers that exactly match the Sylph output file. See the example files provided with the package for guidance, available at `inst/extdata/`.

## Citation

To cite strainspy please use

> Mallawaarachchi S, Tandon K, Rajan N, Marcelino VR, Sandhu S, Bedoui S,
> Ingle DJ, Gunjur A, Tonkin-Hill G (2026). Accurate detection of metagenomic
> strain-level associations using average nucleotide identity with StrainSpy.
> *bioRxiv*. doi:
> [10.64898/2026.08.30.748153](https://doi.org/10.64898/2026.08.30.748153)