---
title: "strainspy Workflow"
author: "Gerry Tonkin-Hill, Sudaraka Mallawaarachchi"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{strainspy Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Why strainspy?

`strainspy` fits strain-level association models from ANI/abundance profiles and metadata.
This guide follows the README workflow and is written as a practical first analysis.

You will walk through the core path most users care about:

1. Load metadata and ANI profiles.
2. Filter to informative strains.
3. Fit a zero-inflated beta model with `glmZiBFit()`.
4. Rank and interpret hits with `top_hits()`.
5. Explore taxonomy-aware summaries and plot-ready outputs.
6. Prepare data for downstream prediction and diversity analyses.

All examples use bundled data so the workflow runs end-to-end.

## Load package and example data

```{r setup, message=FALSE, warning=FALSE}
library(strainspy)

meta_path <- system.file("extdata", "example_metadata.csv.gz", package = "strainspy")
sylph_path <- system.file("extdata", "example_sylph_profile.tsv.gz", package = "strainspy")
taxonomy_path <- system.file("extdata", "example_taxonomy.tsv.gz", package = "strainspy")

meta <- read.csv(meta_path)
meta$Case_status <- factor(meta$Case_status)

taxonomy <- read_taxonomy(taxonomy_path)
se <- read_sylph(sylph_path, meta_data = meta)

# Recommended default: keep strains observed in at least 10% of samples
min_nonzero <- ceiling(0.10 * ncol(se))
se <- filter_by_presence(se, min_nonzero = min_nonzero)

cat("Samples:", ncol(se), "\n")
cat("Retained strains:", nrow(se), "\n")
cat("min_nonzero used:", min_nonzero, "\n")
```

## Choosing key wrapper inputs

If you use the high-level `strainspy()` wrapper, three inputs should be chosen deliberately:

1. `design_formula`: the predictors you want to test.
2. `coef`: which coefficient index from that design to report in `top_hits()`.
3. `min_nonzero`: prevalence threshold (recommended default: `ceiling(0.10 * n_samples)`).

For the design `~ Case_status + Age_at_collection`, `coef = 2` targets `Case_status`.
You can verify term order with `colnames(model.matrix(~ Case_status + Age_at_collection, meta))`.

## Fit the main model (`glmZiBFit`)

```{r fit, message=FALSE, warning=FALSE}
design <- as.formula("~ Case_status + Age_at_collection")

fit <- glmZiBFit(
  se,
  design,
  nthreads = 2L,
  return_vcov = FALSE
)

fit
```

## Find top hits

`top_hits()` is usually the first table you inspect after fitting.
It highlights contigs with strongest phenotype association and provides adjusted p-values.

```{r top-hits, message=FALSE, warning=FALSE}
hits <- top_hits(fit, coef = 2, alpha = 1)
head(hits[, c("Contig_name", "coefficient", "p_value", "p_adjust")])
```

## Add taxonomy context

When taxonomy is available, `hadjust()` summarizes significance across taxonomic levels.

```{r hadjust, message=FALSE, warning=FALSE}
hier_hits <- hadjust(fit, coef = 2, taxonomy = taxonomy)
head(hier_hits[, c("Level", "Model", "Name", "p_adjust")])
```

## Generate plot-ready outputs

The README demonstrates visualization early because it helps with interpretation.
To keep this vignette stable in automated builds, we return data frames (`plot = FALSE`) rather than rendering all figures.
You can set `plot = TRUE` interactively.

```{r plot-data, message=FALSE, warning=FALSE}
plot_volcano(fit, coef = 2)
plot_manhattan(fit, coef = 2, taxonomy = taxonomy)
```

## Diversity and prediction preparation

Beyond hit calling, `strainspy` supports ecological summaries and model-ready tables.

```{r utilities, message=FALSE, warning=FALSE}
rich <- estimate_sample_richness(se, taxonomy)
beta_dist <- estimate_beta_diversity(se, distance_only = TRUE)

pred_df <- prep_for_prediction(
  sy = se,
  outcome = "Case_status",
  contigs = rownames(se)[1:3],
  covariates = "Age_at_collection",
  use_genome_names = FALSE
)

length(rich$species_richness)
class(beta_dist)
dim(pred_df)
```

## Working with other profile inputs

The same downstream workflow can be used with other supported profile readers.

```{r other-inputs, message=FALSE, warning=FALSE}
example_sourmash_path <- system.file("extdata", "example_sourmash.csv.gz", package = "strainspy")
example_metaphlan_path <- system.file("extdata", "metaphlan_merged.tsv.gz", package = "strainspy")

sm <- read_sourmash(example_sourmash_path, meta)
mp <- read_metaphlan(example_metaphlan_path, meta)

class(sm)
class(mp)
```

## Notes

- If taxonomy is not available, sequence-level workflows (`top_hits`, `plot_volcano`) still run.
- For optional features requiring additional packages, see individual function help pages.

## Next steps

After this walkthrough, common next analyses are:

1. Re-fit on the full filtered dataset (`se`) and tighten `alpha` in `top_hits()`.
2. Use taxonomy-informed Manhattan summaries to prioritize clades.
3. Export `prep_for_prediction()` outputs to train external classifiers.

## Session information

```{r session-info}
sessionInfo()
```
